---
title: "Laby_dist"
author: "Sandra"
date: "2025-04-14"
output: html_document
---

# Labyrinthulomycete distribution code

This code is heavily based upon the work of Faith

## Library

```{r}
library(stringr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tibble)
library(readr)
library(readxl)
library(scatterpie)
library(sf)
```

```{r}
tax_levels <- c(
  "domain",
  "supergroup",
  "division",
  "subdivision",
  "class",
  "order",
  "family",
  "genus",
  "species"
)
tax_levels_boot <- str_c(tax_levels, "_boot")
tax_levels_updated <- c("class", "order", "family", "genus", "species")
```

## Importing data (Sandra)

```{r}

asv_new_taxo <- read_excel("raw_data/metabarcode/asv_updated.xlsx")

samples <- read_excel("raw_data/metabarcode/samples.xlsx", guess_max = 100000)

asv_abundance <- read_tsv(file = 'raw_data/metabarcode/metapr2/metapr2_ASVs_selected_abundance_Eukaryota_2025-03-13.tsv')


```

## Importing data (Daniel)

```{r}

# file_path <- function(file) {
#   here::here("data", "metapr2", file)
# }
# 
# asv_new_taxo <- rio::import(file_path("asv_updated.xlsx"))
# samples <- readxl::read_xlsx(file_path("samples.xlsx"), guess_max = 100000)
# asv_abundance <- rio::import(file_path("metapr2_ASVs_selected_abundance_Eukaryota_2025-03-13.tsv.gz"))

```


# Reassigning ASVs sequences


```{r}
asv_abundance <- asv_abundance %>%
  select(!c(domain:species) ) %>%
  left_join(asv_new_taxo)
```


# Data filtration

```{r}

#Removing fungi, metazoa and streptophyta (plants) and making 3 ecosystem types
asv_abundance_filtered <- asv_abundance %>%
  filter(!(subdivision %in% c("Fungi", "Metazoa")))  %>%
  filter(!(division %in% c("Streptophyta"))) %>%
  filter(depth_level == "surface" | depth <= 100) %>%
  mutate(ecosystem_type = case_when(ecosystem %in% c("oceanic", "coastal") ~ ecosystem, TRUE ~ "non-marine" ))
```

# Renormalisation # NEW CODE

```{r}
reads_total_per_sample <- asv_abundance_filtered |>
  group_by(file_code) |>
  summarize(reads_total=sum(n_reads)) |>
  ungroup()

### I had problems with using the rename funciton, so I had to do it like this 
colnames(asv_abundance_filtered)[which(names(asv_abundance_filtered) == "reads_total")] <- "reads_total_old"

asv_abundance_filtered_corrected <- asv_abundance_filtered |>
  left_join(reads_total_per_sample) |>
  mutate (n_reads_pct = n_reads/reads_total*100)

```

# Global distribution of Laby

## Preparing Presence/Absence data - NEW CODE

  P_A     count
  <chr>   <int>
1 absent   1723
2 present  3940

```{r}

laby_PAD <- asv_abundance_filtered_corrected %>% # USE CORRECTED COUNTS
  group_by(file_code, ecosystem_type, latitude, longitude, class) %>%
  dplyr::summarise(total_count = sum(n_reads), .groups = 'drop') %>% # Merging values of the same location
  spread(key = class, value = total_count) %>% # Convert to wide data
  select(file_code, ecosystem_type, latitude, longitude, Labyrinthulomycetes)%>%
  mutate(P_A = case_when((Labyrinthulomycetes > 0) ~ "present",
                         TRUE ~ "absent"))%>%
  as.data.frame()

#Finding how many samples Laby is present in
PA_sum <- laby_PAD %>%
  group_by(P_A) %>%
  summarise(count =n())
```

## Preparing data for dominance map

```{r}

# Filter for class laby - CHANGED TO USE CORRECTED COUNTS
laby_asv_abundance_filter <- filter(asv_abundance_filtered_corrected,class == "Labyrinthulomycetes")

#changing all environmental seqs to one -> Labyrinthulomycetes_X
laby_asv_abundance_filter$order <- gsub("Labyrinthulomycetes_LAB\\d+", 
                                        "Labyrinthulomycetes_X", 
                                        laby_asv_abundance_filter$order)

#Finding the dominant order
dominant_taxon <- laby_asv_abundance_filter %>%
  arrange(file_code, desc(n_reads_pct)) %>%
  group_by(file_code) %>%
  dplyr::slice(1) %>%
  mutate(dominant_taxon = order) %>%
  select(file_code, latitude,longitude ,dominant_taxon,n_reads_pct)

# Merge information on P/A and dominance
dominant_merge <- laby_PAD %>%
  left_join(select(dominant_taxon, dominant_taxon, n_reads_pct) ,   # USE DPLYR JOIN AND NOT MERGE
            by=c('file_code'))

#Finding the number of samples where each order is dominant
taxa_sum <- dominant_merge %>%
  group_by(dominant_taxon) %>%
  summarise(count =n())

#one file_code is >80 percent, seems like an outlier. Removed
dominant_merge <- dominant_merge[dominant_merge$file_code != 'IC2016-127-5', ]
```

## Creating maps

### Main map

```{r}

# Create base map
worldmap <- map_data ("world")# From the tidyverse package

base_world  <- ggplot()+ 
  coord_fixed() +
  xlab("") + ylab("") + 
  geom_polygon(data=worldmap, aes(x=long, y=lat, group=group), 
               fill="grey80")+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_rect(fill = 'white', colour = 'white'), 
        axis.line = element_line(colour = "white"), #legend.position="none",
        axis.ticks=element_blank(), axis.text.x=element_blank(),
        axis.text.y=element_blank())

main_map <- base_world +
  geom_jitter(data = dominant_merge, aes(x = longitude, 
                                           y = latitude,
                                           color = dominant_taxon,
                                           size = n_reads_pct),  # Map size directly to n_reads_pct
               alpha = 0.5) +
  geom_point(data = subset(dominant_merge, is.na(dominant_taxon)), 
             aes(x = longitude, 
                 y = latitude,
                 shape = 'NA'), stroke = 0.1, shape = 4, color = "black") +
  scale_size(range = c(1, 15)) +  # Use same size range for both plot and legend
  scale_color_viridis_d() +
  labs(size = "% of microbial Eukaryotes", color = "Dominant taxon") +
  guides(size = guide_legend()) +  # Keep this simple to maintain scaling
  theme_minimal() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank())
main_map

ggsave("laby_worldmap.pdf",plot = last_plot(),device = pdf, scale = 4, width = 1300, height= 1000, units = "px", dpi=320, path = 'output/metabarcode', limitsize = FALSE)

```

### North Europe

```{r}

# Assuming main_map is your initial map object
eu_map <- main_map +
  coord_sf(
    xlim = c(-22.5, 28),  # Set limits for the x-axis
    ylim = c(50, 80),     # Set limits for the y-axis
    expand = TRUE
  ) +
  theme_minimal() +  # Use a minimal theme to retain axes
  theme(
    legend.position = "none",
    panel.border = element_rect(fill = NA, colour = "black"),  # Add border around the panel
    axis.title.x = element_text(margin = margin(t = 10)),  # Add some margin to the x-axis title
    axis.title.y = element_text(margin = margin(r = 10))   # Add margin to the y-axis title
  ) +
  scale_x_continuous(breaks = seq(-25, 30, 5),  # Specify breaks for the longitude, ensuring 0 is included
                     name = "Longitude") +
  scale_y_continuous(breaks = seq(50, 80, 5),    # Specify breaks for the latitude
                     name = "Latitude")

# Print the map
eu_map

ggsave("laby_neurope.pdf",plot = last_plot(),device = pdf, scale = 2, width = 1500, height= 1300, units = "px", dpi=320, path = 'output/metabarcode', limitsize = FALSE)


```

## Ecosystem distribution

### Data manipulation

```{r}

laby_asv_abundance_filter$species <- gsub("Labyrinthulomycetes_", "", laby_asv_abundance_filter$species) #Shortening the species names to better fit the treemap
laby_asv_abundance_filter$species <- gsub("Thraustochytriidae_", "", laby_asv_abundance_filter$species)


#Finding out how many ASVs is in each ecosystem,
ASVs_tot<- laby_asv_abundance_filter %>%
  group_by(ecosystem_type) %>%
  summarise(tot_asv_count = n_distinct(asv_code))
```

## Scatterpie map

### Scatterpie map peparing

```{r}

# laby_order <- laby_asv_abundance_filter %>%
#     select(file_code,ecosystem_type,latitude,longitude,order,n_reads_pct)

laby_order <- laby_asv_abundance_filter %>%
    select(file_code,ecosystem_type,latitude,longitude,order,n_reads_pct)%>%
    group_by(file_code,ecosystem_type,latitude,longitude,order)%>%
    dplyr::summarise(n_reads_pct = sum(n_reads_pct),.groups = 'drop')%>% # summing values of the same file_code
    arrange(order, desc(n_reads_pct))

laby_order_list <- laby_order |>
  select(order) |>
  distinct() |>
  pull(order)

laby_order <- laby_order |>
    pivot_wider(names_from = order, values_from = n_reads_pct) %>%
    replace(is.na(.), 0) %>% 
    right_join(select(dominant_merge, file_code, ecosystem_type, longitude, latitude, P_A),
               by = join_by(file_code, ecosystem_type, latitude, longitude)) |> # ADDED FOR CROSSES WHERE ABSENT
    as.data.frame() |>
    filter(!is.na(latitude)) # REMOVE MISSING DATA

```

### Building map

```{r}
  ### Plotting scatterpie map of orders in class Laby (surface)
  laby_scatter<-base_world+
    geom_scatterpie(aes(x=longitude, y=latitude),
                    data=filter(laby_order, P_A == "present"), # ADDED SINCE NOW WE HAVE ALSO SAMPLES WHERE ABSENT
                    cols=laby_order_list, # BETTER CODING, USE COLUMN NAMES NOT POSITION
                    color=NA,alpha=0.8,pie_scale = 0.3) +
    geom_point(data = filter(laby_order, P_A == "absent"),
             aes(x = longitude,
                 y = latitude,
                 shape = 'NA'), 
             size = 1,
             stroke = 0.1, 
             shape = 4, 
             color = "black") +
    theme(legend.position = "bottom")+
    guides(fill=guide_legend(title="Order"))+
    facet_wrap(~ecosystem_type)+
    scale_fill_viridis_d() 
  
  laby_scatter
```
```{r}
ggsave("laby_scatter.pdf",plot = last_plot() ,device = pdf, scale = 2, width = 3000, height= 800, units = "px", dpi=320, path = 'output/metabarcode', limitsize = FALSE)
```

## Tree map

### Tree map preparing

```{r}
# Preparing laby treemap data: order and subgroup species (surface)
  laby_order_sub <- laby_asv_abundance_filter %>%
    select(file_code,ecosystem_type,latitude,longitude,order,species,n_reads_pct)%>%
    group_by(ecosystem_type,order,species)%>%  # NO NEED FOR ACROSS
    dplyr::summarise(n_reads_pct = mean (n_reads_pct),.groups = 'drop')%>% # merging values of the same file_code ##THIS IS WHERE DANIEL CHANGES
    drop_na()%>%
    as.data.frame()
  
  #####  FIXED - YOU NEED FIRST TO SUMMARIZE BY CLASS - YOU MUST RESTART FROM ORIGINAL DF
    # Finding average laby % for each variable
  f_labels <- laby_asv_abundance_filter %>%
    select(file_code, ecosystem_type ,class, n_reads_pct)%>%
    group_by(file_code, ecosystem_type, class)%>%  # NO NEED FOR ACROSS
    dplyr::summarise(n_reads_pct = sum (n_reads_pct),.groups = 'drop') %>%
    group_by(ecosystem_type)%>% # NO NEED FOR ACROSS
    dplyr::summarise(avg_reads = mean(n_reads_pct),.groups = 'drop')%>% # merging values of the same file_code
    drop_na()%>%
    as.data.frame()

```

### Building map

```{r}
    
#Creating tree map
laby_tree<- ggplot(laby_order_sub, aes(area = n_reads_pct, 
                                         fill = order,
                                         subgroup = order, 
                                         label = species)) +
    treemapify::geom_treemap() +
    treemapify::geom_treemap_text(colour = "white", place = "centre", grow = FALSE) +
    treemapify::geom_treemap_subgroup_border() +
    treemapify::geom_treemap_subgroup_text(place = "topleft", grow = F, 
                                           alpha = 0.5, colour = "black", 
                                           min.size = 0) +
    theme_bw() +
    # title() +
    scale_fill_viridis_d()+
    facet_wrap(~ecosystem_type)+
    coord_cartesian(xlim = c(50, 350), ylim = c(10, 35), clip = "off")+
    guides(fill = "none") 
    
laby_tree
```
```{r}
ggsave("laby_tree.pdf",plot = last_plot() ,device = pdf, scale = 2, width = 3000, height= 800, units = "px", dpi=320, path = 'output/metabarcode', limitsize = FALSE)
```

