---
title: "metadata_ecosystem"
author: "Sandra"
date: "2025-02-25"
output: html_document
---

# Metadata ecosystem

## Packages

```{r}
library(tidyverse)
library(readxl)
```

## Organizing metadata

The ecosystem metadata has been gathered both from annotations and from reading the source literature of each sequence. Both sequence centroids and hits have metadata, thus the centroid sequences will eventually represent the metadata from both itself and it's hits. Initially, more categories were used to characterize the source in the metadata categorized file, this will be simplified. The following categories are used categorize the ecosystems.

### Oceanic

Marine/Seawater with no more info, Antarctica, Deep sea (below 200 m)

### Coastal

Seagrass, Seaweed, Fjords, Intertidal zone, Lagoon, Benthic diatom film, Bays, Corals, Estuary, Mangroves, Baltic sea

### Freshwater

Freshwater lake, Freshwater river, Bog,

### Terrestrial

Soil, Rice soil, Dung, Turfgrass, Refuse dump

### Other

Artificial structures, Farm reactors

### Note

Type sequences have been manually changed into centroids in the file to make sure their metadata is included.

NOTE: The metadata_categorized file is has sequences added to it which were not part of the clustering, thus the center_seq_name column is not accurate compared to the original file.

## Importing sequences

```{r}
#importing categorized metadata file
meta <- read_csv2("raw_data/metadata_categorized.csv")


#the metadata appears to have some duplicates, these are removed
meta <- meta %>%
  distinct(seq_name, .keep_all = TRUE)

test2 <- anti_join(eco_data_seqs,meta, by = "seq_name")

test <- right_join(meta,eco_data_seqs)

```

## Editing metadata

```{r}

#categorizing into the 5 categorizes
meta <- meta %>%
  mutate(
    pr2_ecosystem = case_when(
      is.na(pr2_ecosystem) ~ NA_character_,
      pr2_ecosystem %in% c("deep_sea", "oceanic") ~ "oceanic",
      pr2_ecosystem %in% c("estuary","coastal", "mangrove") ~ "coastal",
      pr2_ecosystem %in% c("freshwater river", "freshwater lake","limnic","spring","bog") ~ "limnic",
      pr2_ecosystem %in% c("terrestrial") ~ "terrestrial",
                           TRUE ~ "other"
    )
  )

#creating a new df with only the center_seq_name and pr2_ecosystem
meta_eco <- meta %>%
  select(center_seq_name, pr2_ecosystem)

#creating a pivot table
pivot_table <- meta_eco %>%
  count(center_seq_name,pr2_ecosystem) %>%
  pivot_wider(names_from = pr2_ecosystem, values_from = n, values_fill = 0)

#I remove the NA column because it is uneccessary
pivot_table <- pivot_table[, !names(pivot_table) %in% c("NA")]

#renaming the center_seq_name column 
names(pivot_table)[names(pivot_table) == "center_seq_name"] <- "seq_name"

#checking why there are more sequences in the pivot table than the tree
test <- pivot_table %>%
  anti_join(seqs_kept, by = "seq_name")

#gathering information on these sequences, appears that they have been removed
test2 <- left_join(test,sequences)

#checking if there are some sequences of the seqs_kept df that is missing from the pivot table
test3 <- seqs_kept %>%
  anti_join(pivot_table, by = "seq_name")

```

```{r}
#importing excel sheet with all sequences and the labels in the tree
new_taxonomy <- read_excel("raw_data/tree_taxonomy.xlsx")

tree_label <- new_taxonomy %>% select(seq_name,tree_label)

#making a new df with the tree label and pivot table data
meta_eco_label <- left_join(tree_label,pivot_table)




#summing the values in each category 
meta_eco_label_sum <- meta_eco_label %>%
  group_by(tree_label) %>%
  summarise(
    oceanic = sum(oceanic, na.rm = TRUE),
    coastal = sum(coastal, na.rm = TRUE),
    limnic = sum(limnic, na.rm = TRUE),
    terrestrial = sum(terrestrial, na.rm = TRUE),
    other = sum(other, na.rm = TRUE)
  )

#exporting file with data
#write.csv2(meta_eco_label_sum, file = "output/eco_tree_data.csv")
```
